% !TEX TS-program = lualatex
\documentclass[11pt]{article}

% ---------- Packages ----------
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amsthm,amssymb,mathtools,bm}
\usepackage{bbm}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{enumitem}
\usepackage{thmtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=teal,urlcolor=magenta]{hyperref}
\usepackage[nameinlink,capitalize]{cleveref}

% ---------- Bibliography as embedded file ----------
% [removed: embedded refs.bib filecontents]

\usepackage[backend=biber,style=authoryear,doi=true,isbn=false,url=false,giveninits=true,maxcitenames=2]{biblatex}
\addbibresource{refs.bib}

% ---------- Theorem-like environments ----------
\declaretheorem[name=Theorem,numberwithin=section]{theorem}
\declaretheorem[name=Proposition,sibling=theorem]{proposition}
\declaretheorem[name=Lemma,sibling=theorem]{lemma}
\declaretheorem[name=Corollary,sibling=theorem]{corollary}
\declaretheorem[name=Definition,sibling=theorem]{definition}
\declaretheorem[name=Remark,sibling=theorem]{remark}
\declaretheorem[name=Example,sibling=theorem]{example}

% ---------- Macros ----------
\newcommand{\MT}{\mathrm{MT}}
\newcommand{\CT}{\mathrm{CT}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Argmin}{\operatorname*{arg\,min}}
\newcommand{\Argmax}{\operatorname*{arg\,max}}
\newcommand{\inner}[2]{\left\langle #1,\,#2\right\rangle}

% ---------- Title ----------
\title{\textbf{Idiom Projections on Finite Paths and Fisher--Information Transport}}
\author{Jacob Alexander Elliott}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We develop two compositional idioms over finite path spaces and ...ts on the $\CT$ side---the correspondence is local and lossless.

Fisher information (FI) transports \emph{exactly}: $I_{Q(L)}(\theta)\!=\!I_{L}(\theta)$ when $Q$ is an isometry induced by sufficient statistics; under coarse-grainings, FI contracts as expected and the Cram\'er--Rao lower bound (CRLB) degrades monotonically. Quantum FI, via classical embeddings and Petz monotone metrics, mirrors the same data-processing behavior. Category-theoretic scaffolding and enriched structures remain background \parencite{Kelly1982,Elliott2025ModalShell}.
\end{abstract}

% =========================
\section{Setup}\label{sec:setup}
% =========================
Let $G=(V,E)$ be a finite directed graph and $T\in\mathbb{N}$ a ...algebra. Parameters live in an open $\Theta\subset\mathbb{R}^d$.

\paragraph{MT objects.}
An $\MT$ object is $(G,T,\{P_\theta\}_{\theta\in\Theta})$ with $...p_\theta(p)$. The Fisher information matrix (FIM) at $\theta$ is
\begin{equation}
I_P(\theta)
=\E_{P_\theta}\!\big[(\nabla_\theta \ell_\theta)(\nabla_\theta \ell_\theta)^\top\big]
=-\E_{P_\theta}\!\big[\nabla_\theta^2 \ell_\theta\big],
\end{equation}
well-defined since the space is finite \parencite{Fisher1922}.

\paragraph{CT objects.}
Fix $\lambda>0$ (inverse temperature). A $\CT$ object is ...; partition, gradient, and Hessian without path enumeration. The ... maps are enriched; see also \parencite{Giry1982,Fritz2020,Kelly1982} for categorical probability and enriched structures.

% =========================
\section{Information transport}\label{sec:transport}
% =========================
Let $K$ be a Markov kernel (coarse-graining); denote pushforward $Q(F_K L)$ on the $\CT$ side and $Q(L)$ the CT embedding. Then:
\begin{proposition}[Data processing]\label{prop:DP}
$I_{Q(F_K L)}(\theta)\preceq I_{Q(L)}(\theta)$ with equality iff $\nabla_\theta \ell_\theta$ is a.s. measurable w.r.t.\ a sufficient statistic, i.e.\ equality at sufficiency \parencite{Pitman1936,Koopman1936,Darmois1935,Cover2006,Csiszar2011}.
\end{proposition}

\begin{remark}[CRLB transport]
For any locally unbiased estimator $\hat\theta$ under $P_\theta$,
\(
\Cov(\hat\theta)\succeq I_P(\theta)^{-1}.
\)
Under a kernel $K$ the bound degrades monotonically by \cref{prop:DP}. Related invariants are discussed in \textcite{QFI}.
\end{remark}

% =========================
\section{Quantum information}\label{sec:quantum}
% =========================
Embed $Q(L)_\theta$ diagonally as $\rho_\theta=\sum_p Q(L)_\theta(p)\,|p\rangle\!\langle p|$. The quantum Fisher information $F_Q(\rho_\theta)$ satisfies $I_{Q(L)}(\theta)\le F_Q(\rho_\theta)$ with equality characterized by Petz sufficiency \parencite{Petz1986,Helstrom1976,Braunstein1994,Holevo2019,Paris2009}. For any channel $\Lambda$ that coarse-grains outcomes,
$F_Q(\Lambda(\rho_\theta))\le F_Q(\rho_\theta)$, paralleling \cref{prop:DP}. Related invariants are discussed in \textcite{QFI}.
\end{remark}

% =========================
\section{FI examples}\label{sec:examples}
% =========================
... examples, finite sums; proofs rely only on standard manipulations of Markov kernels and exponential families.

% =========================
\section*{Related and external works}
% =========================
\begin{itemize}[leftmargin=2em]
\item \textbf{Information geometry and monotonicity.} The Fisher information induces a Riemannian metric on statistical models; its monotonicity under Markov morphisms and the role of $\alpha$-connections are developed in \textcite{AmariNagaoka2000} and the classic monograph of \textcite{Cencov1981}.
\item \textbf{Classical lower bounds and asymptotics.} The Cram\'er--Rao bound and related inequalities trace to \textcite{Cramer1946} and \textcite{Rao1945}; broader decision-theoretic and asymptotic treatments appear in \textcite{LeCam1986}.
\item \textbf{Comparison of experiments and sufficiency.} For data-processing and sufficiency notions allied to our transport results, see \textcite{Blackwell1951}.
\item \textbf{Bayesian Cram\'er--Rao (van Trees) bound.} For a Bayesian perspective on information bounds under channels, see \textcite{VanTrees1968}.
\end{itemize}

\printbibliography

\end{document}
